---
title: "Data analysis assignment 2"
author: "Andrew Bosbery"
date: "06/02/2020"
output: github_document
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

In this assignment you will work with relational data, i.e. data coming from different data tables that you can combine using keys. Please read ch.13 from R for Data Science before completing this assignment -- https://r4ds.had.co.nz/relational-data.html.

## Read data

We will work with three different tables: household roster from wave 8 (*h_egoalt*), stable characteristics of individuals (*xwavedat*), and household data from wave 8 (*h_hhresp*).

```{r}
library(tidyverse)
# You need to complete the paths to these files on your computer.

Egoalt8 <- read_tsv("C:/Users/Andrew/Documents/University/Year 2/Term 2/Data Analysis 3/Project/data/UKDA-6614-tab/tab/ukhls_w8/h_egoalt.tab")
Stable <- read_tsv("C:/Users/Andrew/Documents/University/Year 2/Term 2/Data Analysis 3/Project/data/UKDA-6614-tab/tab/ukhls_wx/xwavedat.tab")
Hh8 <- read_tsv("C:/Users/Andrew/Documents/University/Year 2/Term 2/Data Analysis 3/Project/data/UKDA-6614-tab/tab/ukhls_w8/h_hhresp.tab")
```

## Filter household roster data (10 points)

The **egoalt8** data table contains data on the kin and other relationships between people in the same household. In each row in this table you will have a pair of individuals in the same household: ego (identified by *pidp*) and alter (identified by *apidp*). *h_relationship_dv* shows the type of relationship between ego and alter. You can check the codes in the Understanding Society codebooks here -- https://www.understandingsociety.ac.uk/documentation/mainstage/dataset-documentation.

First we want to select only pairs of individuals who are husbands and wives or cohabiting partners (codes 1 and 2). For convenience, we also want to keep only the variables *pidp*, *apidp*, *h_hidp* (household identifier), *h_relationship_dv*, *h_esex* (ego's sex), and *h_asex* (alter's sex).

```{r}
Partners8 <- Egoalt8 %>%
        filter(h_relationship_dv == 1 | h_relationship_dv == 2) %>%
        select(pidp, apidp, h_hidp, h_relationship_dv, h_sex, h_asex)
```

Each couple now appears in the data twice: 1) with one partner as ego and the other as alter, 2) the other way round. Now we will only focus on heterosexual couples, and keep one observation per couple with women as egos and men as their alters.

```{r}
Hetero8 <- Partners8 %>%
        # filter out same-sex couples
        filter(h_sex != h_asex) %>%
        # keep only one observation per couple with women as egos
        filter(h_sex == 2 & h_asex == 1)  #h_asex == 1 unnecessary but a good safeguard
```

## Recode data on ethnicity (10 points)

In this assignment we will explore ethnic endogamy, i.e. marriages and partnerships within the same ethnic group. First, let us a create a version of the table with stable individual characteristics with two variables only: *pidp* and *racel_dv* (ethnicity).

```{r}
Stable2 <- Stable %>%
        select(pidp, racel_dv)
```

Let's code missing values on ethnicity (-9) as NA.

```{r}
Stable2 <- Stable2 %>%
        mutate(racel_dv = recode(racel_dv, `-9` = NA_real_))
```

Now let us recode the variable on ethnicity into a new binary variable with the following values: "White" (codes 1 to 4) and "non-White" (all other codes).

```{r}
Stable2 <- Stable2 %>%
        mutate(race = as.character(case_when(
                between(racel_dv, 1, 4) ~ "white",
                racel_dv > 4 ~ "non-white")))
                
        #NAs introduced by coercion
```

## Join data (30 points)

Now we want to join data from the household roster (*Hetero8*) and the data table with ethnicity (*Stable2*). First let us merge in the data on ego's ethnicity. We want to keep all the observations we have in *Hetero8*, but we don't want to add any other individuals from *Stable2*.

```{r}
JoinedEthn <- Hetero8 %>%
        left_join(Stable2, by = "pidp")
        
```

Let us rename the variables for ethnicity to clearly indicate that they refer to egos.

```{r}
JoinedEthn <- JoinedEthn %>%
        rename(egoRacel_dv = racel_dv) %>%
        rename(egoRace = race)
```

Now let us merge in the data on alter's ethnicity. Note that in this case the key variables have different names in two data tables; please refer to the documentation for your join function (or the relevant section from R for Data Science) to check the solution for this problem.

```{r}
JoinedEthn <- JoinedEthn %>%
        left_join(Stable2, by = c("apidp" = "pidp"))
```

Renaming the variables for alters.

```{r}
JoinedEthn <- JoinedEthn %>%
        rename(alterRacel_dv = racel_dv) %>%
        rename(alterRace = race)
```

## Explore probabilities of racial endogamy (20 points)

Let us start by looking at the joint distribution of race (White vs. non-White) of both partners.

```{r}
TableRace <- JoinedEthn %>%
        # filter out observations with missing data
        filter(alterRace != is.na(alterRace) & egoRace != is.na(egoRace)) %>%
        count(egoRace, alterRace)
TableRace
```

Now calculate the following probabilities: 1) for a White woman to have a White partner, 2) for a White woman to have a non-White partner, 3) for a non-White woman to have a White partner, 4) for a non-White woman to have a non-White partner.

Of course, you can simply calculate these numbers manually. However, the code will not be reproducible: if the data change the code will need to be changed, too. Your task is to write reproducible code producing a table with the required four probabilities.


```{r}
TableRace %>%
        # group by ego's race to calculate sums
        group_by(egoRace) %>%
        # create a new variable with the total number of women by race
        mutate(Female_Race_Total = sum(n)
               ) %>%
        # create a new variable with the required probabilities 
        mutate(Alter_Race_Prob = (n/Female_Race_Total) * 100)
```

## Join with household data and calculate mean and median number of children by ethnic group (30 points)

1) Join the individual-level file with the household-level data from wave 8 (specifically, we want the variable for the number of children in the household).
2) Select only couples that are ethnically endogamous (i.e. partners come from the same ethnic group) for the following groups: White British, Indian, and Pakistani.
3) Produce a table showing the mean and median number of children in these households by ethnic group (make sure the table has meaningful labels for ethnic groups, not just numerical codes).
4) Write a short interpretation of your results. What could affect your findings?

```{r}
NKids_Households <- Hh8 %>% 
        select(h_hidp, h_nkids_dv)
#A table with only the identifier and the number of kids per household

JoinedEthn <- JoinedEthn %>%
        left_join(NKids_Households, by = "h_hidp")

#adding the number of kids per household to the couples table


```


```{r}

##Choosing only endogamous couples

Same_Ethnic <- JoinedEthn %>%
        filter(egoRace == alterRace) %>%
        filter(egoRacel_dv == 1 | egoRacel_dv == 9 | egoRacel_dv == 10) #%>%
        #filter(h_nkids_dv.x != is.na(h_nkids_dv.x))

        

```




```{r}
Couple_Table <- Same_Ethnic %>%
        count(egoRacel_dv, h_nkids_dv)
        

        
Couple_Table
        

```





```{r}
Couple_Table %>%
      mutate(weightedkids = n * h_nkids_dv) %>%
      mutate(totalcouplesrace = case_when(
         egoRacel_dv == 1 ~ sum(n[egoRacel_dv == 1]),
         egoRacel_dv == 9 ~ sum(n[egoRacel_dv == 9]),
         egoRacel_dv == 10 ~sum(n[egoRacel_dv == 10])
      )) %>%
       mutate(avergaekids = case_when( 
        egoRacel_dv == 1 ~  sum(weightedkids[egoRacel_dv == 1], na.rm = TRUE)/totalcouplesrace,
        egoRacel_dv == 9 ~  sum(weightedkids[egoRacel_dv == 9], na.rm = TRUE)/totalcouplesrace,
        egoRacel_dv == 10 ~  sum(weightedkids[egoRacel_dv == 10], na.rm = TRUE)/totalcouplesrace
        
        
        
        )
        

       )
                   
     


```

On average white couples have 0.56 childern per household, Indian couples have 0.93 children per household and Pakistani couples have 1.70 children per household. This suggests that white British couples have less children than couples from these two ethnicities and Pakistani couples the most. Socio-economic factors could be a reason for this. This could be that many of the Pakistani and Indian couples have a less secure socio-economic position, they earn less money and do not have much in terms of savings. This could mean that these couples have more children to support them when they retire unlike the white couples who may have a relatively better off income and so do not need as many children to support them in the future. 

Theses findings may be affected by many factors. One of these the is not controlling for age and professional status. This table does not account for the ages of participants or their professional status therefore it is possible this may be a confounding factor. For example, if an unrepresentative amount of the white British couples are young professionals and Indian and Pakistani couples and unrepresentively older with only working than this could explain why white British couples have on average less children than their Indian and Pakistani counterparts. This is because two young professionals may have less time to run a family than an older couple one of whom will be free to look after the kids. If this was accounted for there may have been a smaller difference in the average amount of children at different ages or professional status. 


